Microlensing as a possible probe of the internal structure of quasars

### 2015-02-15 paper1/code.tex
Irshad edits

For the model a it was desirable to work with a caustic geometry as simple as possible. Therefore only two equal point like masses were used as lenses, to produce the magnification map depicted in figure \ref{fig:magnification}. For a description of the parameter set which characterizes the map see (reference). The magnification maps calculated by the microlensing code are in units of Einstein radii, hence they are principally scale free. Physical scale is introduced to them by multiplying the pixels by the correct distance factor. The Einstein angle is given by   

@@

For the next step, the computation of the actual lightcurve, the code was modified to also allow for crescent shaped images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and $R_n$ the inner one. The orientation of the source image with respect to the magnification map is specified by the parameters $a$ and $b$ as the shift of the center of the inner disk from the center of the outer disk in $x$- and $y$-direction respectively. In the original version of the code, gaussian and disk shaped images where already implemented. Those are completely characterized by the single parameter $R_p$. The values of the parameters are specified in pixel units corresponding to the magnification map. Further one needs to specify the start and end point coordinates of the path, which the center of the source image follows through the magnification map (see the depiction in figure \ref{fig:magnification}). Hence the points along this path are specified through the number of timesteps for which the computation of the brightness is to be carried out. Those points correspond to the actual measurement of the brightness of an object in the observational case. For each timestep the two-dimensional convolution of equation \ref{eqn:ft2d} is carried out numerically for the position of the source on the magnification map. \\

>>>>

For the model a it was desirable to work with a caustic geometry as simple as possible. Therefore only two equal point like masses were used as lenses, to produce the magnification map depicted in figure \ref{fig:magnification_map}. For a description of the parameter set which characterizes the map see (reference). The magnification maps calculated by the microlensing code are in units of Einstein radii, hence they are principally scale free. Physical scale is introduced to them by multiplying the pixels by the correct distance factor. The Einstein angle is given by   

@@

For the next step, the computation of the actual lightcurve, the code was modified to also allow for crescent shaped images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and $R_n$ the inner one. The orientation of the source image with respect to the magnification map is specified by the parameters $a$ and $b$ as the shift of the center of the inner disk from the center of the outer disk in $x$- and $y$-direction respectively. In the original version of the code, gaussian and disk shaped images where already implemented. Those are completely characterized by the single parameter $R_p$. The values of the parameters are specified in pixel units corresponding to the magnification map. Further one needs to specify the start and end point coordinates of the path, which the center of the source image follows through the magnification map (see the depiction in figure \ref{fig:magnification_map}). Hence the points along this path are specified through the number of timesteps for which the computation of the brightness is to be carried out. Those points correspond to the actual measurement of the brightness of an object in the observational case. For each timestep the two-dimensional convolution of equation \ref{eqn:ft2d} is carried out numerically for the position of the source on the magnification map. \\

### 2015-02-15 paper1/discussion.tex
Irshad edits

This work presents a toy model to put useful constraints on the interior structure of quasars. To carry out this analysis on real data, one needs two situations: First, a quasar which can be observed with a reasonably good telescope with high signal to noise and second, a caustic crossing event of the quasar. These events are not rare, however, to get an isolated caustic crossing might be. However, as the size of the source is negligible as compared to big caustic structures in the sky, even the single fold caustic crossing is not that rare.

In this toy analysis, we used very ideal data, which might not be the case with real source. So, we propose a more sophisticated analysis of this kind with different combinations of random and systematic noise and selection of data points for future.

The main emphasis of this work was the crescent model parameters, which we show can be well recovered by the maximum likelihood analysis, apart from highly degenerate $a$ and $b$ parameters. The reason of this degeneracy is also very intuitive, if one integrates parallel to the caustic to get the integrated 1D brightness profile, so if the motion of the source is either parallel or perpendicular to the caustic, one of the two parameters ($a$ and $b$) is indistinguishable and in case of angled motion of the source, these two are highly degenerate. This degeneracy can be broken if multiple crossing events of the same quasar is available. So for example, one can use the first crossing to estimate the likelihood of the four parameters and use them as priors in the second crossing and so on. In this way, one can accurately and precisely recover all four parameters.

Further, one can also study another interesting case, where the shape of the source is crescent like but the brightness of the crescent is not uniform but Gaussian.

>>>>

This work presents a toy model to put useful constraints on the interior structure of quasars, particularly the environment geometry of the event horizon. To carry out this analysis on real data, one needs two situations: First, a quasar which can be observed with a reasonably good telescope with high signal to noise ($\sim 150$) and second, a caustic crossing event of the quasar. These events are not rare, however, to get an isolated caustic crossing might be. As the size of the source is negligible as compared to big caustic structures in the sky, even the single fold caustic crossing is not that rare.

In this toy analysis, we used very ideal data, which might not be the case with real source but we contaminated it to match a reasonable signal to noise.

The main emphasis of this work was the crescent model parameters, which we show can be well recovered by the maximum likelihood analysis, apart from highly degenerate $a$ and $b$ parameters. The reason of this degeneracy is also very intuitive, if one integrates parallel to the caustic to get the integrated 1D brightness profile, so if the motion of the source is either parallel or perpendicular to the caustic, one of the two parameters ($a$ and $b$) is indistinguishable and in case of angled motion of the source, these two are highly degenerate. This degeneracy can be broken if multiple crossing events of the same quasar is available. So for example, one can use the first crossing to estimate the likelihood of the four parameters and use them as priors in the second crossing and so on. In this way, one can accurately and precisely recover all four parameters. Using our toy analysis we conclude that it is feasible, in practice, to recover 2D shape of the source using a microlensing event and distinguish between symmetric and asymmetric sources.

### 2015-02-15 paper1/fitting.tex
Irshad edits

\section{Model fitting and likelihood analysis}

@@

In this section, we attempt to mock three datasets, each for crescent, uniform-disk and Gaussian disk case for a given fiducial parameters and contaminate it with systematic and random noise, and fit \& recover the model parameters. So broadly we have 9 cases to discuss, we name these nine cases as different permutations (couplet) of three letters C, D and G for crescent, uniform-disk and Gaussian disk case respectively. The first letter of the couplet represents the fit model and the second represents the dataset. For example, CG represents the case where we fit Gaussian disk data to a crescent model. Similarly, DC respresents the case when crescent data is being fit with uniform-disk model. We explored the parameter space and carried out a likelihood analysis on all three parameter sets using Markov-chain Monte-Carlo (MCMC).

\subsection{Mock datasets and model parameters}
The datasets for each source model is in the form of light curves, the brightness of the source observed at different time. Figure \ref{fig:datafitting} shows the three datasets. These datasets are generated using the magnification map shown in figure \ref{fig:magnification} if the source of the respective model moves from point C to B. The data assumed was very  ideal, including 250 different data points at regular interval of time. We added Gaussian random noise to this dataset of width 10\% of the brightness at the respective time, which makes the signal to noise of about $\sim$150. After adding random noise, the three datasets are looking highly contaminated and no clear intuitive behaviour of fit can be seen.

@@

\begin{equation}
	R_p = 50.0 \ \ \ \rm{for\ Gaussian\ disk\ model}
\end{equation}
\\
where, $R_p$ being the size of the uniform disc and Gaussian disc source model. We also have some nuisance parameters in all these cases: the beginning of the event (I), the end of the event (E), normalisation of the lightcurve (N). We marginalised over all these nuisance parameters. When we kept these three nuisance parameters free to vary in MCMC, it adapts to the amount of data for which the maximum likelihood is most optimal to find out. We talk more about this adaptaion in the next section. Secondly, We try to fit the these datasets for a crossing of the source from a different caustic, in this case, from point A to B in figure \ref{fig:magnification}. This adds a natural systematic in the dataset and the data is no more ideal. With these three characterstics -- random noise, systematic noise and nuisance parameters -- we make the dataset very general and highly contaminated to mock the real observations.

@@

First we explored the parameter space for crescent source model and run its likelihood exploration for three different datasets. Here we have four interesting parameters as given by equation \ref{eqn:cp} and we marginalise over all other nuisance parameters. Three solid curves in top row of figure \ref{fig:mcmc} represents the resulting likelihood of $R_p$, the overall size of the crescent source fitted with three different datasets. We also show results for half light radius $R_{1/2}$ which is calculated using equation \ref{eqn:}.

@@

\item Crescent data: This is the true case for crescent model fitting. The red solid line represents this case in upper panels of \ref{fig:mcmc} as 1D probability density function marginalised over all other parameters. The true value of $R_p$ can be recovered within 2$\sigma$ level with an accuracy up to 10\%. Also the best fit to the data is shown in red line in upper-left panel of figure \ref{fig:datafitting}. It is a very good fit which also represents a reduced $\chi^2$ close to unity (1.04). The orange contour in the lower panels of figure \ref{fig:mcmc} showing the 2D marginalised error ellipses. There is a certain degeneracy in $R_{1/2}$ and $R_n/R_p$ parameters but all the values are well recovered.

@@

Therefore we conclude here that looking at the maximum likelihood one can identity the true dataset and also recover the true values of the parameter.

@@

Lets assume that an ideal telescope observes a quasar for a long time and we get the lightcurve  and one can clearly see the event of caustic crossing in the light curve. Now one wants to extract the true shape of the source. To comment more specifically, lets assume three cases when the true shape of the source is: crescent, uniform disc and Gaussian disc:

@@

At this point, we conclude that maximum likelihood analysis can be well trusted in order to recover true shape and recovering the corresponding parameters. This will provide the accurate recovery of the model parameters, however, for more precise recovery, one rely on precise measurements of the lightcurve.

>>>>

\section{Model fitting and likelihood analysis}\label{sec:fitting}

@@

In this section, we attempt to mock three datasets, each for crescent, uniform-disk and Gaussian disk case for a given fiducial parameters and contaminate it with systematic and random noise, and fit \& recover the model parameters. Broadly we have 9 cases to discuss, we name these nine cases as different permutations (couplet) of three letters C, D and G for Crescent, Uniform-Disk and Gaussian Disk case respectively. The first letter of the couplet represents the fit model and the second represents the dataset. For example, CG represents the case where we fit Gaussian disk data to a crescent model. Similarly, DC respresents the case when crescent data is being fit with uniform-disk model. We explored the parameter space and carried out a likelihood analysis on all three parameter sets using Markov-chain Monte-Carlo (MCMC).

\subsection{Mock datasets and model parameters}
The datasets for each source model is in the form of light curves, the brightness of the source observed at different time. Figure \ref{fig:datafitting} shows the three datasets. These datasets are generated using the magnification map shown in figure \ref{fig:magnification_map} if the source of the respective model moves from point C to B. The data assumed was very  ideal, including 250 different data points at regular interval of time. We added Gaussian random noise to this dataset of width 10\% of the brightness at the respective time, which makes the signal to noise of about $\sim$150. After adding random noise, the three datasets are looking highly contaminated and no clear intuitive behaviour of fit can be seen.

@@

\begin{equation}
	R_p = 50.0 \ \ \ \rm{for\ Gaussian\ disk\ model}
\end{equation}
\\
where, $R_p$ being the size of the uniform disc and Gaussian disc source model. We also have some nuisance parameters in all these cases: the beginning of the event (I), the end of the event (E), normalisation of the lightcurve (N). We marginalised over all these nuisance parameters. When we kept these three nuisance parameters free to vary in MCMC, it adapts to the amount of data for which the maximum likelihood is most optimal to find out. We talk more about this adaptaion in the next section. Secondly, We try to fit the these datasets for a crossing of the source from a different caustic, in this case, from point A to B in figure \ref{fig:magnification_map}. This adds a natural systematic in the dataset and the data is no more ideal. With these three characterstics -- random noise, systematic noise and nuisance parameters -- we make the dataset very general and highly contaminated to mock the real observations.

@@

First we explored the parameter space for crescent source model and run its likelihood exploration for three different datasets. Here we have four interesting parameters as given by equation \ref{eqn:cp} and we marginalise over all other nuisance parameters. Three solid curves in top row of figure \ref{fig:mcmc} represents the resulting likelihood of $R_p$, the overall size of the crescent source fitted with three different datasets. We also show results for half light radius $R_{1/2}$.

@@

\item Crescent data: This is the true case for crescent model fitting. The red solid line represents this case in upper panels of figure \ref{fig:mcmc} as 1D probability density function marginalised over all other parameters. The true value of $R_p$ can be recovered within 2$\sigma$ level with an accuracy up to 10\%. Also the best fit to the data is shown in red line in upper-left panel of figure \ref{fig:datafitting}. It is a very good fit which also represents a reduced $\chi^2$ close to unity (1.04). The orange contour in the lower panels of figure \ref{fig:mcmc} showing the 2D marginalised error ellipses. There is a certain degeneracy in $R_{1/2}$ and $R_n/R_p$ parameters but all the values are well recovered.

@@

Therefore we conclude here that looking at the maximum likelihood one can identify the true dataset and also recover the true values of the parameters.

@@

Lets assume that an ideal telescope observes a quasar for a long time and we get the lightcurve  and one can clearly see the event of caustic crossing in the light curve. Now one wants to extract the true shape of the source. To comment more specifically, lets assume three cases when the true shape of the source is: crescent, uniform disc or Gaussian disc:

@@

Therefor, we conclude that maximum likelihood analysis can be well trusted in order to recover true shape and recovering the corresponding parameters. This will provide the accurate recovery of the model parameters, however, for more precise recovery, one rely on precise measurements of the lightcurve.

### 2015-02-15 paper1/intro.tex
Irshad edits

>>>>

In section \ref{sec:fitting}, we carried out a likelihood analysis for the parameterisation of the different source shapes and study the degeneracies involved amongst them. To do so, we mocked the datasets corressponding to each source shape and contaminate them with different noise components. Finally we run MCMC to explore the parameter space of the source shape and some nuisance parameters. The motivation of this part is to explore the possibility if using a simple toy model is it possible to differentiate between symmetric and asymmetric sources based on maximum likelihood analysis.
